---
title: "Ortholog filtering"
author: "Igor Trujnara"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(scales)
library(reshape2)
```

In this document we will analyze the results of filtering the orthologs by score. We will use a slightly larger sample of 111 queries because the folder I used contained more results and I couldn't be bothered to filter.

We load the data.

```{r load_data}
setwd("out_1k/orthologs/score_table/")
hits_1 <- read.csv("hits_1.csv")
hits_2 <- read.csv("hits_2.csv")
hits_3 <- read.csv("hits_3.csv")
```

We add an extra column for convenience.

```{r merge_hits}
hits_1$threshold <- 1
hits_2$threshold <- 2
hits_3$threshold <- 3
hits <- rbind(hits_1, hits_2, hits_3)
```

Histograms, you know the drill.

```{r plot_total}
ggplot(hits) +
  aes(x = total) +
  geom_histogram(bins=100) +
  facet_wrap(~threshold, scale = "free_x") +
  labs(x = "Number of orthologs", y = "Count", title = "Ortholog counts by threshold")
```

```{r plot_total}
ggplot(hits) +
  aes(x = total) +
  geom_histogram(bins=100) +
  facet_wrap(~threshold, scale = "free_x") +
  labs(x = "Number of orthologs", y = "Count", title = "Ortholog counts by threshold")
```

```{r plot_total_overlaid_log}
ggplot(hits) +
  aes(x = total, fill = as.factor(threshold)) +
  geom_histogram(bins=100, alpha = 0.6) +
  scale_x_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  labs(x = "Number of orthologs", y = "Count",
       fill = "Minimum score", title = "Ortholog counts by threshold",
       caption = "Zero values have been removed") +
  theme_minimal()
```

```{r plot_total_density_log}
total_thr_plot <- ggplot(hits %>% transform(total = total + 0.01)) +
  aes(x = total, fill = as.factor(threshold)) +
  geom_density(alpha = 0.6) +
  scale_x_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  scale_fill_brewer(palette = "Accent") +
  labs(x = "Number of orthologs (log10 scale)", y = "Density",
       fill = "Minimum score", title = "Total ortholog counts by threshold") +
  theme_minimal()

total_thr_plot
```

```{r plot_total_box}
ggplot(hits) +
  aes(x = as.factor(threshold), y = total) +
  geom_boxplot() +
  labs(title = "Hit counts by threshold", x = "Number of orthologs", y = "Count")
```

This is unreadable. I have two ideas on how to fix it.

```{r plot_total_box_ylim}
ggplot(hits) +
  aes(x = as.factor(threshold), y = total) +
  geom_boxplot() +
  ylim(-10, 2000) +
  labs(title = "Hit counts by threshold", x = "Number of orthologs", y = "Count")
```

```{r plot_total_box_log}
ggplot(hits %>% filter(threshold < 3) %>% transform(total = total + 0.01)) +
  aes(x = as.factor(threshold), y = total, fill = as.factor(threshold)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  scale_fill_brewer(palette = "Accent") +
  labs(title = "Ortholog counts by threshold", fill = "Minimum score", x = "", y = "Count", 
       caption = "Zero values have been removed") +
  theme_minimal()
```

Zooms now.

```{r total_zoom_1}
ggplot(hits_1) +
  aes(x = total) +
  geom_histogram(bins = 100) +
  xlim(-10,1000) +
  labs(x = "Number of orthologs", y = "Count", title = "Score >= 1")
```

```{r total_zoom_2}
ggplot(hits_1) +
  aes(x = total) +
  geom_histogram(bins = 100) +
  xlim(-10,200) +
  labs(x = "Number of orthologs", y = "Count", title = "Score >= 1")
```

```{r total_zoom_3}
ggplot(hits_2) +
  aes(x = total) +
  geom_histogram(bins = 26) +
  xlim(-3,25) +
  labs(x = "Number of orthologs", y = "Count", title = "Score >= 2")
```

Now by database.

```{r oma_plots}
ggplot(hits) +
  aes(x = OMA) +
  geom_histogram(bins = 100) +
  facet_wrap(~threshold, scales = "free_x") +
  labs(title = "OMA", x = "OMA hits", y = "Count")
```

```{r oma_alt}
ggplot(hits %>% filter(threshold < 3) %>% transform(OMA = OMA + 0.01)) +
  aes(x = OMA, fill = as.factor(threshold)) +
  geom_density(alpha = 0.6) +
  scale_x_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  labs(x = "Number of OMA hits", y = "Density", fill = "Minimum support",
       title = "OMA hits by threshold")
```

```{r comparison_density}
hits_long <- hits %>%
  filter(threshold < 3) %>%
  melt(id.vars = c("id", "threshold", "total")) %>%
  transform(value = value + 0.01)

comparison_thr <- ggplot(hits_long) +
  aes(x = value, fill = as.factor(threshold)) +
  geom_density(alpha = 0.6) +
  scale_x_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  scale_fill_brewer(palette = "Accent") +
  facet_wrap(~variable) +
  labs(x = "Number of orthologs (log10 scale)", y = "Density", fill = "Minimum support") +
  theme_minimal()

comparison_thr
```

```{r panther_plots}
ggplot(hits) +
  aes(x = PANTHER) +
  geom_histogram(bins = 100) +
  facet_wrap(~threshold, scales = "free_x") +
  labs(title = "PANTHER", x = "PANTHER hits", y = "Count")
```

```{r eggnog_plots}
ggplot(hits) +
  aes(x = EggNOG) +
  geom_histogram(bins = 100) +
  facet_wrap(~threshold, scales = "free_x") +
  labs(title = "EggNOG", x = "EggNOG hits", y = "Count")
```

```{r compare_plot}
ggplot(hits %>% filter(threshold < 3) %>% melt(id.vars = c("id", "threshold", "total"))) +
  aes(x = as.factor(threshold), y = value, fill = as.factor(threshold)) +
  facet_wrap(~variable) +
  geom_boxplot() +
  scale_y_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  labs(x = "", y = "Number of orthologs", fill = "Minimum support") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Observations:

- there is no consensus
- there is a major drop when we filter by score >= 2
- Peer Bork should go to FAIR data jail
