---
title: "Ortholog data"
author: "Igor Trujnara"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggpubr)
library(scales)
```

This is a preview based on 100 random proteins.

We load the data:
```{r read_data}
setwd("out_1k")
stats <- read.csv("orthologs/stats/aggregated_stats.csv")
hits <- read.csv("orthologs/stats/aggregated_hits.csv")
```

Let's look at the stats first. We will examine the empirical distributions.
```{r summary_max}
summary(stats$percent_max)
```

This is all zeros. Disappointed but not surprised. We don't even have full ID mapping offline.

```{r summary_privates}
summary(stats$percent_privates)
```

This is pretty high, although fortunately we see signs of some agreement. Note that the zeros mean nothing was found.

```{r summary_goodness}
summary(stats$goodness)
```

Keeping in mind that the real range is from 0.33 to 1.0, those values are really low. Same note for zeros.

Now let's make some plots. We skip percentage of consensus as it's boring.
```{r privates_plot}
privates_plot <- ggplot(stats) +
  aes(x = percent_privates) +
  geom_histogram(bins = 50) +
  labs(x = "Percentage of privates", y = "Count") +
  theme_light()
privates_plot
```

```{r privates_plot_but_nicer}
privates_plot <- ggplot(stats) +
  aes(x = percent_privates) +
  geom_density(fill = "darkorange", alpha = 0.6) +
  labs(x = "Percentage of privates", y = "Density",
       caption = "Zero values indicate lack of data; no true zeros") +
  theme_minimal()
privates_plot
```

```{r goodness_plot}
goodness_plot <- ggplot(stats) +
  aes(x = goodness) +
  geom_density(fill = "darkolivegreen1", alpha = 0.6) +
  geom_vline(xintercept = 0.33, color = "firebrick") +
  xlim(-0.1, 1) +
  labs(x = "Goodness", y = "Density",
       caption = "Zero values indicate lack of data") +
  theme_minimal()
goodness_plot
```

```{r combine_stat_plots}
ggarrange(privates_plot, goodness_plot, labels = c("A", "B"))
```

Barring zeros, the values are indeed low. Not that as there is no consensus, this plot is nearly the mirror of the privates plot.

Let us now look at hit counts. Empirical statistics first.
```{r summary_hits}
summary(hits$total)
```

There is huge variation in hit counts. There are zeros, but some groups hit tens of thousands. A median of 382 does not sound bad. Let's look at each database.

```{r summry_oma}
summary(hits$OMA)
```

OMA yields a bit less hits than the average (although, as Toni claims, good ones). Some spikes seem to exist though.

```{r summary_panther}
summary(hits$PANTHER)
```

This shows a similar picture to OMA.

```{r summary_eggnog}
summary(hits$EggNOG)
```

Surprising. There are lots of zeros, but if EggNOG finds something, it finds a lot.

Let's make plots now.
```{r plot_total}
plot_total <- ggplot(hits) +
  aes(x = total) +
  geom_histogram(bins = 100) +
  theme_light()
plot_total
```

Let's zoom on the small values.

```{r plot_zoom}
plot_total_zoom <- ggplot(hits) +
  aes(x = total) +
  geom_histogram(bins = 100) +
  xlim(-100,3000) +
  theme_light()
plot_total_zoom
```

```{r plot_total_cooler}
total_plot <- ggplot(hits %>% transform(total = total + 0.01)) +
  aes(x = total) +
  geom_density(alpha = 0.6, fill = "darkorchid1") +
  scale_x_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
  labs(title = "Total ortholog counts", x = "Number of orthologs (log10 scale)", y = "Density") +
  theme_minimal()

total_plot
```

The numbers are mostly small, but zeros are infrequent. There is a good number of high-hit queries as well.

Now the databases.

```{r plot_oma}
plot_oma <- ggplot(hits) +
  aes(x = OMA) +
  geom_histogram(bins = 100) +
  theme_light()
plot_oma
```

More zeros here. However, a vast majority of queries got something. Very high values are rare.

```{r plot_panther}
plot_panther <- ggplot(hits) +
  aes(x = PANTHER) +
  geom_histogram(bins = 100) +
  theme_light()
plot_panther
```

Interesting. Very few zeros, but a lot of small values. Extremely few above 500.

```{r plot_eggnog}
plot_eggnog <- ggplot(hits) +
  aes(x = EggNOG) +
  geom_histogram(bins = 100) +
  theme_light()
plot_eggnog
```

This is a big number. Zoom time.

```{r plot_eggnog_zoom}
plot_eggnog_zoom <- ggplot(hits) +
  aes(x = EggNOG) +
  geom_histogram(bins = 100) +
  xlim(-100, 3000) +
  theme_light()
plot_eggnog_zoom
```

That is a lot of zeros. And no particular trend in the values above.

Let's compare the distributions.

```{r comparison}
comparison <- hits %>%
  select(-total) %>%
  transform(OMA = OMA + 0.01, PANTHER = PANTHER + 0.01, EggNOG = EggNOG + 0.01) %>%
  pivot_longer(!id, names_to = "source", values_to = "count") %>%
  ggplot() +
    aes(x = source, y = count, fill = source) +
    ylim(-100,3000) +
    geom_violin(alpha = 0.6) +
    scale_fill_viridis_d() +
    scale_y_continuous(trans = "log10", labels = label_comma(scale_cut = cut_short_scale())) +
    labs(title = "Ortholog counts by database", x = "Density by source", y = "Number of orthologs (log10 scale)") +
    theme_minimal() +
    theme(legend.position = "none")
comparison
```

The distributions look similar, with the differences described above. EggNOG has a very visible tail.

